Ecological management of stochastic systems with long transients

### 2020-07-07 manuscript/manuscript.Rmd
first pass revisions done

Systems with alternative stable states [@Scheffer2001; @Carpenter] which can exhibit tipping points [@Scheffer2009; @Carpenter] are of great theoretical interest and pose unique challenges to ecological management [].

@@

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossman1980; @Hastings2018].  A decrease in the parameter $a$ could be thought to model the weakened ability of the trees to respond to attack, arising from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

@@

To properly understand the consequences transient dynamics have for management, we must first be more explicit about both the management objectives and potential management actions.  For simplicity, I will imagine a manager has the option to apply a direct mortality to the beetle larvae prior to recruitment, e.g. by spraying at some given intensity. Thus the action space is defined by some measure of spraying effort between 0% and 100%.  The management objectives are defined by a utility function, which trades off the costs of control with the costs of damage due to the outbreak.  Here I assume linearly increasing cost of control (fixed price per unit spraying). Below an endemic population size $X_0$ I assume damage (e.g. no tree mortality) is minimal, and then increases quadratically with increasing beetle density; that is, the instantaneous utility of state with beetle density $X_t$ and taking action $A_t$ is

@@

@@

\begin{equation}
V_t(X_t) = \max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)
\end{equation}

where the evolution of $X_t$ is determined by Eq \eqref{model}. Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.  By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985; @Marescot2013].

@@

```{r figure3, fig.cap="Optimal policy functions under each model. Parameter values for utility function used here are $c = 1$, $d = 0.5$, $X_0 = 50$, $\\delta = 0.999$, parameters for the transition models are as given in Fig 1."}
## Fig 3

>>>>

Systems with alternative stable states which can exhibit critical transitions are of great theoretical interest [@Scheffer2001; @Ratajczak2018] and pose unique challenges to ecological management [@Folke2004; @Fischer2009; @Polasky2011].  Most of this work has focused on critical transitions driven by saddle-node bifurcations [@Boettiger2013], which produce a blue-sky transition between alternative stable states and exhibit hysteresis which can frustrate efforts to restore the pre-transition state. Less attention has been given to the dynamics which immediately follow the saddle node bifurcation, in which the effects of the anhilated stable state may still be felt as a "Ghost Attractor," as introduced in the molecular and neuroscience literature [e.g. @Grossberg1980; @Deco2012], and more recently in the ecological literature [@Hastings2018].  As such, despite the relative simplicity of the model, Ghost Attractor dynamics can provide important insights to our understanding of long transients, alternative stable state dynamics,and critical transitions, and the consequences such phenomena have for ecological management.

@@

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossberg1980; @Hastings2018].  A decrease in the parameter $a$ could be thought to model the weakened ability of the trees to respond to attack, arising from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

@@

To properly understand the consequences transient dynamics have for management, we must first be more explicit about both the management objectives and potential management actions.  For simplicity, I will imagine a manager has the option to apply a direct mortality to the beetle larvae prior to recruitment, e.g. by spraying at some given intensity. Thus the action space is defined by some measure of spraying effort between 0% and 100%.  The management objectives are defined by a utility function, which trades off the costs of control with the costs of damage due to the outbreak.  Here I assume linearly increasing cost of control (fixed price per unit spraying). Below an endemic population size $X_0$ I assume damage (e.g. no tree mortality) is minimal, and then increases quadratically with increasing beetle density; that is, the instantaneous utility of state with beetle density $X_t$ and taking action $A_t$ (the "utility function") is

@@

Where `d` is the coefficient representing the damage (in terms of lost utility), and `c` is the coefficient representing the cost to providing control.  Note that while it is common to express utility in strictly monetary terms (i.e. damage to the forest expressed in dollars lost, vs dollars spent to control the outbreak), it is by no means an assumption of the approach that utility must maximize monetary reward.  In particular, different managers may assign different utility to either forest state or the control effort (i.e. differ in their values of `d` and `c`), and thus reach different management strategies, both of which are 'optimal' relative to their separate utilities.  See @Walters1986, @Clark1990, @Fischer2009, and @Marescot2013 for further discussion of utility maximization in ecological management.

@@

\begin{equation}
V_t(X_t) = \max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t) \label{bellman}
\end{equation}

where the evolution of $X_t$ is determined by Eq \eqref{model}. Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.  By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the Hamilton-Jacobi-Bellman recursion, Eq \eqref{bellman}, using Stochastic Dynamic Programming (SDP) [@Mangel1985; @Marescot2013].

@@

```{r figure3, fig.cap="Optimal policy functions under each model. A policy function indicates the optimal action (y-axis) that should be taken given each possible state (x-axis) of the system. The policy function arises as the solution to the Hamilton-Jacobi-Bellman recursion, Eq (4), e.g. as found by stochastic dynamic programming. Parameter values for utility function used here are $c = 1$, $d = 0.5$, $X_0 = 50$, $\\delta = 0.999$, parameters for the transition models are as given in Fig 1."}
## Fig 3
